Co-moving stars in \textsl{Gaia DR1}

### 2016-10-05 paper/ms.tex
add a bunch of words [ci skip]

\todo{Some bullshit about binary stars...}

@@

The primary data set used in this \documentname\ is the Tycho-Gaia Astrometric
Solution (TGAS), released as a part of Data Release 1 of the Gaia mission
(\citealt{many}). TGAS contains astrometric measurements (sky position,
parallax, and proper motions) and associated covariance matrices for a large
fraction of the \project{Tycho-2} catalog (\citealt{tycho2}) with median
astrometric precision equivalent to the \project{Hipparcos} catalog ($\approx
0.3~{\rm mas}$; \citealt{hipparcos}). In terms of parallax signal-to-noise
($\snr = \varpi/\sigma_\varpi$), the TGAS catalog contains 1240114 stars
with $\snr > 4$ and 619618 stars with $\snr > 8$. The increase in precision over
Tycho-2 and increase in volume over Hipparcos makes the TGAS catalog a powerful
data set to search for widely-separated, co-moving stars.

In the sections that follow, we construct a statistical model that use the TGAS
data and associated covariances to handle the non-trivial uncertainties [...].
Here, we instead visualize simple quantities constructed directly from the
astrometric measurements to emphasize the quality of the data. For each star
with $\snr > 8$, we use the reported parallax and parallax error to compute a
distance corrected for the Lutz-Kelker bias (\citealt{lutzkelker}):
\begin{equation}
  d = 1000 \, \left[\frac{\varpi}{2} \, \left(1 + \sqrt{1 - \frac{16}{\snr^2}} \right) \right]^{-1} \, {\rm pc}
\end{equation}

@@

For a pair of stars $(i,j)$, we want to ratio the fully marginalized likelihood
(FML) of the stars having the same 3-space velocity (hypothesis 1,
$\mathcal{L}_1$) over the FML of the stars having different 3-space velocities
(hypothesis 2, $\mathcal{L}_2$). When this ratio is large, the pair is a
candidate binary star. For \gaia\ DR1, we only observe parallax, $\varpi$, and
two proper motion components, $\propm = (\mu_\alpha, \mu_\delta)$ for each star.
Where there is overlap with \rave\ or \apogee, some number of observed radial
velocities, $v_r$, with observed variances $\sigma^2_{v_r}$ (independent of the
\gaia\ data) may also be available. We assume that the uncertainties in these
observed quantites are known with covariance matrix $\matrx{C}$. The likelihoods
$\mathcal{L}_1, \mathcal{L}_2$ are marginalized over true 3-space velocity and
distance for each star in the pair.
\begin{align}
  \mathcal{L}_1 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \bs{v} \,
    L_i(\bs{v}, d_i) \,
    L_j(\bs{v}, d_j) \,
    p(\bs{v}) \, p(d_i) \, p(d_j) \\
  \mathcal{L}_2 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \bs{v}_i \, \dd^3 \bs{v}_j \,
    L_i(\bs{v}_i, d_i) \,
    L_j(\bs{v}_j, d_j) \,
    p(\bs{v}_i) \, p(\bs{v}_j) \, p(d_i) \, p(d_j) \\
\end{align}
where
\begin{align}
  L(\bs{v}, d) &=
    \left[\det\left(\frac{\matrx{C}^{-1}}{2\pi}\right)\right]^{1/2} \,
    \exp \left[ -\frac{1}{2} \transp{\left(\bs{x} - \bs{x}_\theta \right)} \,
    \matrx{C}^{-1} \,
    \left(\bs{x} - \bs{x}_\theta \right) \right] \\
  \bs{x} &= \transp{\left(\begin{array}{c c c c} \varpi & \mu_\alpha &
    \mu_\delta & v_r \end{array}\right)} \\
  \bs{x}_\theta &= \transp{\left(\begin{array}{c c c c} \frac{1}{d} & \frac{v_\alpha}{d} &
    \frac{v_\delta}{d} & \tilde{v}_r \end{array}\right)}
\end{align}
\todo{Holy fuck that vector is UGLY}
We assume Gaussian priors on the true velocity components and use a \todo{which
one will we use?} distance prior \citep{Astraatmadja:2016} on heliocentric
distance
\begin{align}
  p(\bs{v}) &= \normal(0,25)~\kms \\
  p(d) &= \frac{r^2}{2\,L^3} \, \exp\left(-r/L\right)
\end{align}

For hypothesis 1, the integral over velocity is non-trivial because the same
``true'' 3-space velocity vector projects into tangential velocity components
differently depending on sky position when the sky separation is appreciable
($v_\alpha = v_\alpha(\bs{v}, \alpha, \delta)$). For hypothesis 2, the integral
can be split into two the product of two simpler integrals $\mathcal{L}_2 = Q_i
\, Q_j$ where
\begin{equation}
  Q = \int \, \dd d \, \dd^3 \bs{v} \, L(\bs{v}, d) \, p(\bs{v}) \, p(d)
\end{equation}

@@

\appendix
\section{Some mathematics relevant to the above}

>>>>

Two stars moving with similar full-space velocity vectors (``co-moving stars'')
are either remnants of a dispersing stellar system or are a widely-separated
binary star system. In either case, co-moving stars are of great interest for
studying chemical uniformity and star formation processes, the destruction of
star clusters and dwarf galaxies, and the tidal field and small-scale properties
of the mass distribution in the Milky Way. To date, thousands of candidate
co-moving star pairs have been identified by searching for stars with common
proper motions (\citealt{Luyten:1979,Poveda:1994,Allen:2000,Gould:2003,
Chaname:2004,Lepine:2007,Alonso-Floriano:2015}).
% Cite:
% https://ui.adsabs.harvard.edu/#abs/2012ApJ...757..170A/abstract
% https://ui.adsabs.harvard.edu/#abs/2012AJ....144..102T/abstract
% https://ui.adsabs.harvard.edu/#abs/2014ApJ...790..158A/abstract

% Two stars moving with full-space velocity vector differences $\lesssim 1~\kms$
% with orbital semi-major axes $a \gtrsim 10^-3~{\rm pc}$.

Useful for:
- star cluster disruption and dispersal
- chemical abundance comparison (no interaction)
- dynamics

@@

The primary data set used in this \documentname\ is the Tycho-Gaia Astrometric
Solution (TGAS), released as a part of Data Release 1 of the Gaia mission
(\citealt{many}). TGAS contains astrometric measurements (sky position,
parallax, and proper motions) and associated covariance matrices for a large
fraction of the \project{Tycho-2} catalog (\citealt{tycho2}) with median
astrometric precision equivalent to the \project{Hipparcos} catalog ($\approx
0.3~{\rm mas}$; \citealt{hipparcos}). In terms of parallax signal-to-noise
($\snr = \varpi/\sigma_\varpi$), the TGAS catalog contains 1240114 stars with
$\snr > 4$, 619618 stars with $\snr > 8$, and 197105 stars with $\snr > 16$. The
increase in precision over Tycho-2 and increase in volume over Hipparcos makes
the TGAS catalog a powerful data set to search for widely-separated, co-moving
stars.

In the sections that follow, we construct a statistical model that propagates
the non-trivial uncertainties in the data to our beliefs about the likelihood
that a given pair are co-moving. Here, we instead compute quantities directly
from the astrometric measurements to visualize and emphasize the quality of the
data. For each star with $\snr > 16$, we use the reported parallax and parallax
error to compute a distance corrected for the Lutz-Kelker bias
(\citealt{lutzkelker}):
\begin{equation}
  d_{\rm LK} = 1000 \, \left[\frac{\varpi}{2} \,
    \left(1 + \sqrt{1 - \frac{16}{\snr^2}} \right) \right]^{-1} \, {\rm pc}
\end{equation}
We use this distance and the sky positions of each source to compute 3D
Cartesian positions, $\boldsymbol{x}$, for all stars and find the nearest
neighbor to each source in 3-space using the KDTree implementation in
\project{scikit-learn} (\citealt{scikit-learn}). We remove duplicates from
this sample and are left with 138962 pairs of stars. For all pairs, we then
compute the tangential velocity difference
\begin{equation}
  |\Delta \bs{v}_{\rm t}| = d_{\rm LK} \,
                            \sqrt{(\mu_{\alpha,1}-\mu_{\alpha,2})^2 +
                                  (\mu_{\delta,1}-\mu_{\delta,2})^2}
\end{equation}
\figname~\ref{fig:dv-sep} shows these quantities plotted for all pairs that pass
the above signal-to-noise cut. Pairs with small physical separation also tend to
have small velocity differences. Intriguingly, there appear to be an abundance
of pairs with small velocity separation at separations $|\delta\bs{x}| \gtrsim
1~{\rm pc}$.

\begin{figure*}[p]
\begin{center}
\includegraphics[width=\textwidth]{figures/dv-sep.pdf}
\end{center}
\caption{%
Difference in tangential velocity vs. physical separation for pairs of stars
selected to be nearest neighbors in separation. Dashed line shows zero velocity
difference. Note the points with small velocity difference: these are likely
widely-separated binary stars.
\label{fig:dv-sep}}
\end{figure*}

\todo{More naive visualizations, or save for the full sample?}

@@

For a given pair of stars $(i,j)$, we compute the fully marginalized likelihood
(FML) that the pair has the same 3-space velocity (hypothesis 1,
$\mathcal{L}_1$) and the FML of the stars having different 3-space velocities
(hypothesis 2, $\mathcal{L}_2$). We use the ratio of these FMLs as a scalar for
selecting candidate widely-separated binary stars. \gaia\ \dr\ only contains
parallax, $\varpi$, and two proper motion components, $\propm = (\mu_\alpha,
\mu_\delta)$ for each star, so for the majority of sources we only have
information about the tangential velocities of the stars. Where there is overlap
with \rave\ or \apogee, some number of observed radial velocities, $v_r$, with
observed variances $\sigma^2_{v_r}$ (independent of the \gaia\ data) may also be
available. We assume that the uncertainties in these observed quantites are
known with covariance matrix $\matrx{C}$. The likelihood functions $L_1, L_2$
are marginalized over true 3-space velocity and distance for each star in the
pair.
\begin{align}
  \mathcal{L}_1 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \bs{v} \,
    L_i(\bs{v}, d_i) \,
    L_j(\bs{v}, d_j) \,
    p(\bs{v}) \, p(d_i \given \varpi_i) \, p(d_j \given \varpi_j) \\
  \mathcal{L}_2 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \bs{v}_i \, \dd^3 \bs{v}_j \,
    L_i(\bs{v}_i, d_i) \,
    L_j(\bs{v}_j, d_j) \,
    p(\bs{v}_i) \, p(\bs{v}_j) \, p(d_i \given \varpi_i) \, p(d_j \given \varpi_j) \label{eq:hyp2}
\end{align}
where
\begin{align}
  L(\bs{v}, d) &=
    \left[\det\left(\frac{\matrx{C}^{-1}}{2\pi}\right)\right]^{1/2} \,
    \exp \left[ -\frac{1}{2} \transp{\left(\bs{x} - \bs{x}_\theta \right)} \,
    \matrx{C}^{-1} \,
    \left(\bs{x} - \bs{x}_\theta \right) \right] \\
  \bs{x} &= \transp{(\begin{array}{c c c} \mu_\alpha &
    \mu_\delta & v_r \end{array})} \\
  \bs{x}_\theta &= \transp{\left(\begin{array}{c c c} d^{-1}\,v_\alpha &
    d^{-1}\,v_\delta & \tilde{v}_r \end{array}\right)}
\end{align}
and we compute the posterior pdfs for the true distances, $p(d\given\varpi)$,
using a uniform space density prior \citep{Astraatmadja:2016} on heliocentric
distance
\begin{equation}
  p(d) \propto d^2
\end{equation}
We assume Gaussian priors on the true velocity components
\begin{equation}
  p(\bs{v}) = \normal(0,25)~\kms
\end{equation}

For hypothesis 1, the integral over velocity is non-trivial because the same
``true'' 3-space velocity vector projects into tangential velocity components
differently depending on sky position ($v_\alpha = v_\alpha(\bs{v}, \alpha,
\delta)$). The marginalization integral for hypothesis 2 is simpler because the
integral in \eqname~\ref{eq:hyp2} can be split into the product of two simpler
integrals $\mathcal{L}_2 = Q_i \, Q_j$ where
\begin{equation}
  Q = \int \, \dd d \, \dd^3 \bs{v} \, L(\bs{v}, d) \, p(\bs{v}) \, p(d\given\varpi)
\end{equation}
In either case, the marginalization over true 3-space velocity can be performed
analytically; we derive the relevant expressions in
\sectionname~\ref{sec:appendix}. After marginalizing over velocity, the
likelihood integrands only depend on distance; we numerically compute the
integrals over the true distances to each star in a pair, $d_1,d_2$, using Monte
Carlo integration with $K$ samples from the distance posterior pdfs:
\begin{equation}
  \int \, \dd d \, \tilde{L}(d) \, p(d\given\varpi) \approx
    \frac{1}{K} \, \sum_k^K \, \tilde{L}(d_k)
\end{equation}
Through experimentation, we have found that $K=128$ samples are sufficient for
estimating the above integrals for stars with a wide range in parallax
signal-to-noise.

@@

\appendix

\section{Some mathematics relevant to the above} \label{sec:appendix}

\todo{clean this up}
